---
title: "p8105_hw3_rs4338"
author: "Rebecca Shyu"
date: "2024-10-08"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(haven)
```

## Problem 0:

* Create a public GitHub repo + local R Project: p8105_hw3_rs4338
* Create a single .Rmd file named p8105_hw3_rs4338.Rmd that renders to github_document
* Create a subdirectory (data) to store the local data files, and use relative paths to access these data files
* Submit a link to your repo via Courseworks: https://github.com/rysgpd/p8105_hw3_rs4338

```{r problem2_importdata, message=FALSE}

nhanes_demo = read_csv("data/nhanes/nhanes_covar.csv", 
                       skip = 4, 
                       col_types = cols(
                            education = "factor")) %>% 
  janitor::clean_names() %>% 
  drop_na() %>% 
  filter(
    age >= 21
  ) %>% 
  mutate(
    sex = ifelse(sex == 1, "Male", "Female")
  )

nhanes_acc_data = read_csv("data/nhanes/nhanes_accel.csv") %>% 
    janitor::clean_names()

```

```{r pressure, echo=FALSE}
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
