---
title: "p8105_hw3_rs4338"
author: "Rebecca Shyu"
date: "2024-10-08"
output: github_document
always_allow_html: true
---

```{r setup, include=FALSE}
library(tidyverse)
library(haven)
library(kableExtra)
library(leaflet)

knitr::opts_chunk$set(
	echo = TRUE,
	warning = FALSE,
  fig.width = 8,
  fig.asp = .6,
  out.width = "100%",
	dpi=300
)

theme_set(theme_minimal() + theme(legend.position = "bottom"))

options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)

scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete = scale_fill_viridis_d

```

## Problem 0:

* Create a public GitHub repo + local R Project: p8105_hw3_rs4338
* Create a single .Rmd file named p8105_hw3_rs4338.Rmd that renders to github_document
* Create a subdirectory (data) to store the local data files, and use relative paths to access these data files
* Submit a link to your repo via Courseworks: https://github.com/rysgpd/p8105_hw3_rs4338

## Problem 1:

* Extra practice on own.

## Problem 2:

```{r problem2_importdata, message=FALSE}

nhanes_demo = read_csv("data/nhanes/nhanes_covar.csv", 
                       skip = 4) %>% 
  janitor::clean_names() %>% 
  drop_na() %>% 
  filter(
    age >= 21
  ) %>% 
  mutate(
    sex = ifelse(sex == 1, "Male", "Female"),
    education = 
      case_match(
        education,
        1 ~ "Less than high school",
        2 ~ "High school equivalent",
        3 ~ "More than high school"
      ),
    education = factor(education, levels = c("Less than high school", "High school equivalent", "More than high school"))
  ) 

nhanes_acc_data = read_csv("data/nhanes/nhanes_accel.csv", show_col_types = FALSE) %>%
  janitor::clean_names() %>% 
  distinct() %>% 
  pivot_longer(
    cols = min1:min1440,
    names_to = "minute",
    values_to = "mims_data",
    names_prefix = "min"
  ) %>% 
  mutate(minute = as.numeric(minute))

# join the demographic and accelerometer datasets
nhanes = inner_join(nhanes_demo, nhanes_acc_data, by="seqn")
# there are 22 patients who don't have any demographic data
```

Reader friendly table for the number of men and women in each education category, and create a visualization of the age distributions for men and women in each education category.

* I decided to use a density plot to visualize the age distributions for men and women in each education category because the histogram was not easily interpretable and rigid. I overlapped the density plots for men and women to easily compare them. 
* It appears that the distributions across all educational levels for men are bi-modal while women are unimodal. 
* For the less than high school education level, it leaned to more older populations for both men and women.
* For the high school equivalent education level, it looked like a steady incline for women and it peaked around age 70. For men, it was pretty consistent throughout and then declined after age 60. 
* For the more than high school education level, women had up to 0.025 density around age 30, which shows more education among younger women. Similarly, but not as high in comparison to women, younger men were higher than older men. 

```{r prob2_genderedu}

nhanes %>% 
  distinct(seqn, .keep_all = TRUE) %>%
  group_by(education, sex) %>% 
  count() %>% 
  arrange(education) %>% 
  knitr::kable() %>% 
  collapse_rows(columns = 1)

nhanes %>% 
  distinct(seqn, .keep_all = TRUE) %>%
  group_by(education, sex) %>% 
  ggplot(aes(x = age, fill = sex)) +
  geom_density(alpha = .3) +
  facet_grid(. ~ education) +
  labs(
    title = "Density Plots of Ages Across Genders & Education Levels" 
    ) +
  theme(plot.title = element_text(hjust = 0.5))

```

Comments about the Total Activity Across Ages, comparing men and women and for each education level:

* For all education levels and both sexes, there is a generally downward trend of total activity (minutes) measured. There are a few exceptions that have slight peaks around ages 60 (less than high school for both, more than high school for men) or 70 (high school equivalent).
* For high school equivalent and more than high school, women are almost always higher than men. For less than high school, younger women have higher activity until age 40, then the men have higher total activity. 
* The trend lines start at the highest for less than high school, but ends at the lowest also for the less than high school level. More than high school seems to have less variability in general, but has the most extreme outliers. 


```{r prob2_total_activity, message=FALSE}

nhanes = 
  nhanes %>% 
  group_by(seqn) %>% 
  mutate(
    total_activity = sum(mims_data)
  )

nhanes %>% 
  distinct(seqn, .keep_all = TRUE) %>% 
  ggplot(aes(x = age, y = total_activity, color = sex)) +
  geom_point(alpha = .8) +
  geom_smooth(se = FALSE) +
  facet_grid(. ~ education) +
  labs(
    title = "Scatterplots of Ages x Total Activity Across Education Levels" 
    ) +
  theme(plot.title = element_text(hjust = 0.5))
```

Accelerometer Data Over 24 Hours for each education level and across sexes.

```{r}

nhanes %>% 
  ggplot(aes(x = minute, y = mims_data, color = sex)) +
  geom_point(size = 0.001, alpha = 0.2) +
  geom_smooth(se = FALSE, linewidth=0.5) +
  facet_grid(. ~ education) +
  labs(
    title = "24 Hour Accelerometer Data Across Education Levels by Sex" 
    ) +
  theme(plot.title = element_text(hjust = 0.5))

```

